LHC abort gap monitor electronics upgrade

### 2024-05-02 paper.tex
Update on Overleaf.

\begin{abstract}
    The LHC Abort Gap Monitor (AGM) is part of the LHC machine protection
    system (MPS) and is designed to measure the particle population in a \SI{3}{\micro\second}
    wide region known as the "abort gap." This region needs to be kept empty to
    ensure safe beam dumps. The AGM captures the synchrotron light generated in
    the visible part of the spectra and converts it into an electric signal.
    This signal is then processed by an acquisition system and can trigger the
    ‘abort gap cleaning’ process. The current AGM, which has been in operation
    since 2010, uses an analogue integrator ASIC and a 40 MHz
    analogue-to-digital (ADC) converter to provide the particle population
    information. However, this solution is now considered obsolete and is being
    replaced by a digital signal processing approach. Working directly in the
    digital domain not only offers more scalability but also better determinism
    and reliability. This work presents the new technical solution for the
    acquisition chain, compares the characteristics of both implementations,
    and showcases recent measurements conducted on the LHC ion and proton
    beams.
\end{abstract}

\section{Introduction}
During the flat-top phase, the LHC ring stores up to \SI{350}{MJ} of beam energy,
spread across 2808 bunches with an average beam transverse size under
\SI{1}{mm}. This high energy density can melt metallic surfaces and quench
superconducting magnets, posing a risk to the machine~\cite{LHC_report}. 
To safely extract the beam from the vacuum compartments, a series of kicker
magnets belonging to the beam dump system is used. These magnets take \SI{3}{\micro\second} to
ramp up their magnetic field~\cite{beam_dump_system}. This defines the size of
an abort gap (AG) in the bunch train. Due to system imperfections, a~residual amount of particles
populate this gap. The average AG population is in order of $10^{9}$ particles. The
population in a single bunch is four orders of magnitude higher in comparison.
When the beam dump is initiated, the particles in AG do not receive full
extraction force. That causes deflection towards the vacuum chamber. When a~certain energy accumulates in the AG, magnets can quench. Thus,
a~continuous monitoring of the AG population is necessary, which is the purpose
of the AGM~\cite{particles_in_ag}.\\
This paper reviews the upgrade of the current AGM system in the LHC since 2010,
focusing on system architecture, analog electronics chain, temperature
dependence, timing instabilities, and a~comparison with the upgraded system.

\section{System architecture}
The AGM system comprises of three main parts. The first part is the optical front-end
(OF), which includes~a pre-amplifier positioned beneath the vacuum chamber. The
second part is a data acquisition system (DAQ) deployed on an FPGA platform
located outside the radiation area. The third part involves a~control software (SW) that
retrieves, interprets, and presents the measured data, facilitating interaction
with the instrument. The system architecture is illustrated in
Fig.~\ref{fig:bsra_chain}, with each part labeled accordingly.
\begin{figure}[!tbh]
    \centering
    \includegraphics[width=\columnwidth]{bsra_chain.pdf}
    \caption{BSRA Chain.}
    \label{fig:bsra_chain}
\end{figure} 
The diagram on the left depicts the layout of the LHC, with beam 1 represented
in red and beam 2 in blue. It includes all four experiments (Alice, Atlas,
LHCb, and CMS). The AGM is located at Point 4 (indicated by the yellow
square), close to the CMS experiment. The Timing, Trigger, and Control system
(TTC) is shown as a purple rectangle in the center of the LHC ring. This system
receives RF clocking and broadcasts it into the network (TTC
distribution)~\cite{ttc_distribution}. The AGM extracts bunch clock information
(corresponding to the position of bunches) and turn clock information (marking
the start of a~turn) from the LHC timing block messages.\\
Beam intensity is measured using the synchrotron light (SL) radiation, as indicated by the red line directed towards the Multi-channel photomultiplier
(MCP-PMT) in the optical front-end (OF). The beam traverses first optical
filters, which increases the system dynamic range. Subsequently, the signal is
transmitted over a \SI{100}{m} cable. To compensate for losses in the cable,
the signal is amplified beforehand. Then it proceeds to the Data Acquisition
(DAQ) system, which measures \SI{6.4}{\micro\second} window and integrates the
intervals of \SI{100}{\nano\second}. This creates set of 64 values published to
the SW every \SI{10}{\micro\second}.

\subsection{Working principle}
The measurement period is determined by the MCP-PMT used, which is the gated
HAMAMATSU R5916U-50 with a duty cycle of \SI{1}{\%}. With a gate pulse duration
of \SI{3}{\micro\second} and one turn cycle length of \SI{89}{\micro\second},
measurements are taken every $4^{\text{th}}$ clock cycle (\SI{356}{\micro\second}),
reducing the duty cycle to \SI{0.84}{\%}.\\
The system captures 64 values with a granularity of \SI{100}{\nano\second}. The
MCP-PMT gate duration is \SI{3}{\micro\second} providing an additional degree of
freedom in the system control. Each \SI{100}{\nano\second} value is averaged in an
extra Exponential Moving Average (EMA) and provided to the software every \SI{100}{\micro\second}. This
time is sufficient to react to changes in AG. The population can change from
1.6 to \SI{25}{\second} due to different
effects~\cite{high_sensitivity_measurement}.

@@

\subsection{Constraints and limitations}
The present AGM exhibits artifacts in the measurements. Signals outside the
gating range do not exhibit the characteristics of white noise, attributed to
the lhcb2002b ASIC. The gain of the integrators lacks linearity, featuring distinct
offsets and dead times. Consequently, the system requires correction for
these effects at each integrator, leading to uncertainties in the outcomes and
necessitating intricate calibration procedures.\\
The system relies on an outdated FPGA platform, rendering its development
software incompatible with the latest operating systems. Additionally, the
fixed amplifier configuration contributes to reduced resolution for signals of
smaller magnitudes.

>>>>

\begin{abstract}
    The LHC Abort Gap Monitor (AGM) is part of the LHC machine protection
    system (MPS) and is designed to measure the particle population in a \SI{3}{\micro\second}
    wide region known as the "abort gap". This region needs to be kept empty to
    ensure safe beam dumps. The AGM captures the synchrotron light generated in
    the visible part of the spectrum and converts it into an electric signal.
    This signal is then processed by an acquisition system and can trigger the
    ‘abort gap cleaning’ process. The current AGM, which has been in operation
    since 2010, uses an analogue integrator ASIC and a 40 MHz
    analogue-to-digital (ADC) converter to provide the particle population
    information. However, this solution is now considered obsolete and is being
    replaced by a digital signal processing approach. Working directly in the
    digital domain not only offers more scalability but also better determinism
    and reliability. This work presents the new technical solution for the
    acquisition chain, compares the characteristics of both implementations,
    and showcases recent measurements conducted on the LHC ion and proton
    beams.
\end{abstract}

\section{Introduction}
During the flat-top phase, the LHC ring stores up to \SI{350}{MJ} of beam energy,
spread across 2808 bunches with an average beam transverse size under
\SI{1}{mm}. This high energy density can melt metallic surfaces and quench
superconducting magnets, posing a risk to the machine~\cite{LHC_report}. 
To safely extract the beam from the vacuum compartments, a series of kicker
magnets belonging to the beam dump system is used. These magnets take \SI{3}{\micro\second} to
ramp up their magnetic field~\cite{beam_dump_system}. This defines the size of
an abort gap (AG) in the bunch train. Due to system imperfections, a residual amount of particles populate this gap.  The typical AG population levels are two to three orders of magnitude lower in comparison to a single LHC bunch.
When the beam dump is initiated, the particles in AG do not receive full
extraction force. That causes deflection towards the vacuum chamber. When a certain energy accumulates in the AG, magnets can quench. Thus,
a~continuous monitoring of the AG population is necessary, which is the purpose
of the AGM \cite{particles_in_ag}.\\
This paper reviews the upgrade of the current AGM system in the LHC since 2010,
focusing on system architecture, analog electronics chain, temperature
dependence, timing instabilities, and a comparison with the upgraded system.

\section{System architecture}
The AGM system comprises of three main parts. The first part is the optical front-end
(OF), which includes~a pre-amplifier positioned beneath the vacuum chamber. The
second part is a data acquisition system (DAQ) deployed on an FPGA platform
located outside the radiation area. The third part involves a~control software (SW) that
retrieves, interprets, and publishes the measured data, facilitating interaction
with the instrument. The system architecture is illustrated in
Fig.~\ref{fig:bsra_chain}, with each part labeled accordingly.
\begin{figure}[!tbh]
    \centering
    \includegraphics[width=\columnwidth]{bsra_chain.pdf}
    \caption{BSRA Chain.}
    \label{fig:bsra_chain}
\end{figure} 
The diagram on the left depicts the layout of the LHC, with beam 1 represented
in red and beam 2 in blue. It includes all four experiments (Alice, Atlas,
LHCb, and CMS). The AGM is located at Point 4 (indicated by the yellow
square), next to the CMS experiment. The Timing, Trigger, and Control system
(TTC) is shown as a purple rectangle in the center of the LHC ring. This system
receives RF clocking and broadcasts it into the network (TTC
distribution)~\cite{ttc_distribution}. The AGM extracts bunch clock information
(corresponding to the position of bunches) and turn clock information (marking
the start of a~turn) from the LHC timing block messages.\\
Beam intensity is measured using the synchrotron light (SL) radiation, as indicated by the red line directed towards the Multi-channel photomultiplier
(MCP-PMT) in the optical front-end (OF). The beam traverses first Neutral Density optical filters, which increases the system dynamic range. Subsequently, the signal is
transmitted over a \SI{100}{m}-long cable. To compensate for losses in the cable,
the signal is amplified beforehand. Then it proceeds to the Data Acquisition
(DAQ) system, which measures \SI{6.4}{\micro\second} window and integrates the
intervals of \SI{100}{\nano\second}. This creates set of 64 values published to
the SW every \SI{10}{\micro\second}.

\subsection{Working principle}
The measurement period is determined by the MCP-PMT used, which is the gated
HAMAMATSU R5916U-50 with a maximum duty cycle of \SI{1}{\%}. With a gate pulse duration
of \SI{3}{\micro\second} and one turn cycle length of \SI{89}{\micro\second},
measurements are taken every $4^{\text{th}}$ clock cycle (\SI{356}{\micro\second}),
reducing the duty cycle to \SI{0.84}{\%}.\\
The system captures 64 values with a granularity of \SI{100}{\nano\second}. The
MCP-PMT gate duration is \SI{3}{\micro\second} providing an additional degree of
freedom in the system control. Each \SI{100}{\nano\second} value is averaged in an
extra Exponential Moving Average (EMA) and provided to the software every \SI{100}{\micro\second}. This
time is sufficient to react to changes in AG. The population can change from
1.6 to \SI{25}{\second} due to different
effects~\cite{high_sensitivity_measurement}.

@@

\subsection{Constraints and limitations}
The present AGM exhibits artifacts in the measurements. Signals outside the gating range do not exhibit the characteristics of white noise, attributed to the lhcb2002b ASIC. The gain of the integrators lacks linearity, featuring distinct offsets and dead times. Consequently, the system requires correction for these effects at each integrator, leading to uncertainties in the outcomes and necessitating intricate calibration procedures. In addition, the amplifier bandwidth (50 kHz - 1,5 GHz) makes it necessary to implement a correction for the missing DC- 50 kHz band to correctly measure AG profiles needed during LHC commissioning phases.\\
The system relies on an outdated FPGA platform, rendering its development
software incompatible with the latest operating systems. Additionally, the
fixed amplifier configuration contributes to reduced resolution for signals of
smaller magnitudes.
